Conservation decision-making under long transients

### 2019-08-12 slides/boettiger.Rmd

---
layout: true
background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
background-size: 120%
background-position: center
class: left, top, inverse

@@

```{r det}
det <- det_sim(may,p) %>% mutate(reps=1) %>% filter(t < 3000)
det  %>%
  ggplot(aes(t, x)) + 
  geom_line(lwd = 2, col="white")
```
---
layout: true
background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
background-size: 150%
class: center, top, inverse
background-position: left

@@

---
background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .85)), url(img/royalsociety/reef-shark.jpg)
background-size: 120%
class: top, inverse
background-position: center
layout: true

@@

```{r posteriors, fig.width=13, fig.height=6}
label <- c("no switch", "switch", "no ghost")

df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
  map_dfr(read_csv, .id = "timeseries") %>% 
  mutate(timeseries = label[as.integer(timeseries)])

@@

p1 <- df %>% 
  ggplot((aes(value, fill=timeseries))) + 
  geom_histogram(alpha=0.9) + 
  geom_vline(aes(xintercept=0.27, col = "no switch"), col= viridis::viridis_pal()(3)[1], lwd=2) + 
  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = viridis::viridis_pal()(3)[2], lwd=2) + 
  ggplot2::scale_fill_viridis_d()

@@

---
layout: true
background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/forest-fire.jpg)
background-size: 120%
class: center, top, inverse

# Descision Theory

@@

### Management under uncertainty

@@

@@

@@

<!--
- Lizzie Wolkovich
- Henry Scharf
-->

>>>>

---
layout: true
background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
background-size: 120%
background-position: center
class: left, top, inverse

@@

```{r det}
det <- det_sim(may,p) %>% mutate(reps=1) %>% filter(t < 3000)
det  %>%
  ggplot(aes(t, x)) + 
  geom_line(lwd = 2, col="white")
```
---
layout: true
background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
background-size: 150%
class: center, top, inverse
background-position: left

@@

---
background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
background-size: 120%
class: top, inverse
background-position: center
layout: true

@@

```{r posteriors, fig.width=13, fig.height=6}
label <- c("no switch", "switch", "no ghost")
pal <- viridis::viridis_pal()(3)
df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
  map_dfr(read_csv, .id = "timeseries") %>% 
  mutate(timeseries = label[as.integer(timeseries)])

@@

p1 <- df %>% 
  ggplot((aes(value, fill=timeseries))) + 
  geom_histogram(alpha=0.9) + 
  geom_vline(aes(xintercept=0.27, col = "no switch"), col= pal[1], lwd=2) + 
  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = pal[2], lwd=2) + 
  ggplot2::scale_fill_viridis_d()

@@

---
layout: true
background-image: linear-gradient(to right, rgba(30, 30, 30, .8), rgba(30, 30, 30, .9)), url(img/royalsociety/forest-fire.jpg)
background-size: 120%
class: center, top, inverse

@@

# Descision Theory
### Management under uncertainty
.pull-left[

## Social

### Utility fn

]
.pull-right[

## Ecological

### Transition fn

]

@@

# Example systems

.pull-left[
## Outbreak

![](img/flickr-vsmoothe-pine-beetle.jpg)

]
.pull-right[

- manage at acceptable/endemic level?
- supress entirely?

]

```{r}
library(MDPtoolbox)
library(sarsop) # remotes::install_github("boettiger-lab/sarsop")
library(tidyverse) # for plotting
library(mdplearning)

damage <- 0.05
control <- 1
reward_fn <- function(x,h) - damage * x ^ 2 - control * h
discount <- 0.98

states <- seq(0,2, length=200)
actions <- states
observations <- states
sigma_g <- 0.05
sigma_m <- 0.0

r <- 0.8
K <- 1.53
q <- 2
b <- .2
eps <- states[2]/10
Tmax <- 100

may <- function(a){  
  function(x, h){ # May
    # x <- pmax(x - h, 0)   # harvest then recruit   
    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps - h
  }
}

possible_a <- seq(.25, .34, by = 0.005)
true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
believe_a <- 0.31 ## believe there's a strong attractor
true_i <- which.min(abs(possible_a - true_a))
```

**True model: weak ghost attractor**

```{r true_a}
f <- may(true_a)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + 
  geom_line(col="white") +
  geom_point(col="white") + 
  geom_hline(aes(yintercept = 0),col="white")
```

**Incorrect belief of bistable system**

```{r a29}
f <- may(0.29)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + 
  geom_line(col="white") +
  geom_point(col="white") + 
  geom_hline(aes(yintercept = 0),col="white")
```

**Incorrect belief of stronger bistable system**

```{r a31}
f <- may(believe_a)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + 
  geom_line(col="white") +
  geom_point(col="white") + 
  geom_hline(aes(yintercept = 0), col="white")
```

---
class: center, middle

# Example management scenarios

```{r}
m_true <- fisheries_matrices(states, actions, observations, reward_fn, 
                        may(true_a), sigma_g, sigma_m, noise = "lognormal")

m <- fisheries_matrices(states, actions, observations, reward_fn, 
                        may(believe_a), sigma_g, sigma_m, noise = "lognormal")
soln <- mdp_value_iteration(m$transition, m$reward, discount)
opt_soln <- mdp_value_iteration(m_true$transition, m_true$reward, discount)

**no Intervention**

```{r no_intervention}
x0 <- which.min(abs(states - 0.1))
no_policy <- numeric(length(states)) + 1

df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
                   policy = no_policy, x0 = x0, Tmax = 100)

df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) +
  geom_point(col="white")+
  geom_line(aes(time, action), col="grey", lwd=2) + ylim(0,1.4)
```

**Mistaken ghost for alternate stable state**:

```{r mistaken_ghost}
x0 <- which.min(abs(states - 0.1))
no_policy <- numeric(length(states)) + 1

df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
                   policy = soln$policy, x0 = x0, Tmax = 100)

df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) +
  geom_point(col="white")+
  geom_line(aes(time, action), col="grey", lwd=2) + ylim(0,1.4)
```

## Adaptive managment: revising beliefs about the process

### Prior beliefs

```{r priors}
prior <- dnorm(possible_a, 0.29, 0.01)
prior <- prior / sum(prior)
data.frame(a = possible_a, probability = prior) %>%
  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill="white") +
  geom_vline(aes(xintercept = true_a), col="white", lwd=1, lty=2) 
```

### Learning too late

### More agnostic prior beliefs

```{r priors_wide}
prior <- dnorm(possible_a, 0.29, 0.01)
prior <- prior / sum(prior)
data.frame(a = possible_a, probability = prior) %>%
  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill="white") +
  geom_vline(aes(xintercept = true_a), col="white", lwd=1, lty=2) 
```

### Less learning, more caution

@@

.pull-left[

Image Credits:

![](img/royalsociety.jpg) Photo Competition 
]
.pull-right[

@@

- Lizzie Wolkovich
- Henry Scharf

]
